### 2015-07-24 doc/thesis/glossary.tex
Fixing glossary definitions and adding page referenes to them

\begin{description}
\item[bottom type] It is a type that is subtype of all types.

\item[closed table type] It is a table type that does not allow adding fields.

\item[coercion] It is a relation that allows converting values from one type to values
of another type without error.

\item[consistency] It is a relation that gradual typing uses to control the insertion of
run-time checks between dynamically typed and statically typed code.

\item[consistent-subtyping] It is a relation that combines consistency and subtyping.

\item[contravariant] Subtyping is contravariant when it reverses the subtyping order,
that is, it orders types from a more generic type to a more specific type.

\item[covariant] Subtyping is covariant when it preserves the subtyping order,
that is, it orders types from a more specific type to a more general type.

\item[depth subtyping] It allows changing the type of individual fields of a table type.

\item[dynamic type] It is a type that allows combining dynamic and static typing in the same code.
It is neither the bottom nor the top type in the subtyping relation, but
a subtype only of itself.
Gradual typing uses the dynamic type along with the consistency relation to
identify where run-time casts should be inserted to prevent that dynamically
typed code violates statically typed code.

\item[filter type] It is a type that allows Typed Lua to discriminate the type of local variables
inside control flow statements.

\item[fixed table type] It is a table type that does not allow adding fields. It also does not allow
width subtyping to make single inheritance safe.

\item[flow typing] It is a combination of static typing and flow analysis.

\item[free assigned variable] It is a free variable that appears in an assignment.

\item[gradual type system] It is a type system that uses the consistency relation to check the interaction
among the dynamic type and other types.

\item[gradual typing] It is an approach that uses a gradual type system to allow static and dynamic
typing in the same code, but inserting run-time checks between statistically
typed and dynamically typed code.

\item[invariant] Subtyping is invariant when it does not allow ordering types, that is,
it is a way that express type equality through subtyping.

\item[metatable] It is a Lua table that allows changing its behavior.

\item[nominal type system] It is a type system that uses name declarations to define the equivalence among types.

\item[open table type] It is a table type that allows adding fields, and usually refers to table types that
have aliases.

\item[optional type system] It is a type system that allows combining static and dynamic typing in the same
language, but without affecting the run-time semantics.

\item[projection environment] It is a type environment that Typed Lua uses to assign projection variables to
second-level types.

\item[projection type] It is a type that allows Typed Lua to discriminate the type of local variables
that have a dependency relation.

\item[prototype object] It is an object that works like a class, that is, it is a prototype for creating
new objects of a given class.

\item[self-like delegation] It is a technique to implement inheritance in dynamically typed languages.

\item[sound type system] It is a type system that does not type check all programs that contain a type error.

\item[structural type system] It is a type system that uses the type structure to define the equivalence among types.

\item[table refinement] It is an operation that allows changing a table type to include new fields or
to specialize existing fields.

\item[top type] It is a type that is supertype of all types.

\item[type environment] It is an environment that assigns variable names to types.

\item[type tag] In dynamically typed languages a type tag describes the type of a value during
run-time.

\item[unique table type] It is a table type that allows adding fields, and it is also the table type
that describe the type of the table constructor.

\item[unsound type system] It is a type system that type checks certain programs that contain type errors.

\item[userdata] It is a way to define new types in an application or library that is written in C.

\item[vararg expression] It is a Lua expression that can result in an arbitrary number of values.

\item[variadic function] It is a Lua function that can receive an arbitrary number of arguments.

\item[variance] It is the way types are ordered.

\item[width subtyping] It allows adding more fields to a table type.

>>>>

\begin{description}
\item[bottom type] A type that is subtype of all types.

\item[closed table type] A table type that does not provide any guarantees
about keys with types not listed in the table type.
See complete definition in page \pageref{def:tabletype}.

\item[coercion] A relation that allows converting values from one type to
values of another type without error.

\item[consistency] A relation used by gradual typing to check the interaction
between the dynamic type and other types.
See complete definition in page \pageref{def:consistency}.

\item[consistent-subtyping] A relation that combines consistency and subtyping.
See complete definition in page \pageref{def:consistent-subtyping}.

\item[contravariant] A part of the type constructor is contravariant when
it reverses the subtyping order.

\item[covariant] A part of the type constructor is covariant when
it preserves the subtyping order.

\item[depth subtyping] An operation that allows variance in the type of
record fields.

\item[dynamic type] A type used by gradual typing to denote unknown values.
See complete definition in page \pageref{def:dynamictype}.

\item[filter type] A type used by Typed Lua to discriminate the type of
local variables inside control flow statements.
See complete definition in page \pageref{def:filtertype}.

\item[fixed table type] A table type which guarantees that there are no
keys with a type that is not one of its key types, and that can have
any number of \emph{fixed} or \emph{closed} references point to it.
See complete definition in page \pageref{def:tabletype}.

\item[flow typing] An approach that combines static typing and flow analysis to
allow variables to have different types at different parts of the program.

\item[free assigned variable] A free variable that appears in an assignment.

\item[gradual type system] A type system that uses the consistency relation
instead of type equality to perform static type checking.
See complete definition in page \pageref{sec:gradual}.

\item[gradual typing] An approach that uses a gradual type system to allow
static and dynamic typing in the same code, but inserting run-time checks
between statically typed and dynamically typed code.
See complete definition in page \pageref{sec:gradual}.

\item[invariant] A part of the type constructor is invariant when it forbids variance.
It is also a way to define type equality through subtyping.

\item[metatable] A Lua table that allows changing the behavior of other tables
it is attached to.

\item[nominal type system] A type system that uses the type names to check the
compatibility among them.

\item[open table type] A table type which guarantees that there are no
keys with a type that is not one of its key types, and that only have
\emph{closed} references point to it.
See complete definition in page \pageref{def:tabletype}.

\item[optional type system] A type system that allows combining static and
dynamic typing in the same language, but without affecting the run-time semantics.
See complete definition in page \pageref{sec:optional}.

\item[projection environment] An environment used by Typed Lua to handle unions of
second-level types that are bound to projection types.

\item[projection type] A type used by Typed Lua to discriminate the type of local
variables that have a dependency relation.
See complete definition in page \pageref{def:projectiontype}.

\item[prototype object] An object that works like a class, that is, it is an object from
which other objects inherit its attributes.

\item[self-like delegation] A technique to implement inheritance in dynamically typed
languages through prototype objects.

\item[sound type system] A type system that does not type check all programs that contain a type error.

\item[structural type system] A type system that uses type structures to check the compatibility among them.

\item[table refinement] An operation from Typed Lua that allows programmers to change a table type
to include new fields or to specialize existing fields.
See complete definition in page \pageref{sec:refinement}.

\item[top type] A type that is supertype of all types.

\item[type environment] An environment used by Typed Lua to assign variable names to first-level types.

\item[type tag] A tag that describes the type of a value during run-time in dynamically
typed languages.

\item[unique table type] A table type which guarantees that there are no
keys with a type that is not one of its key types, and that do not have
any references point to it.
See complete definition in page \pageref{def:tabletype}.

\item[unsound type system] A type system that type checks certain programs that contain type errors.

\item[userdata] A Lua data type that allows Lua to hold values from applications
or libraries that are written in C.

\item[vararg expression] A Lua expression that can result in an arbitrary number of values.

\item[variadic function] A function that can receive an arbitrary number of arguments.

\item[variance] A way to define the subtyping order between the components
of a type constructor.

\item[width subtyping] An operation that allows adding fields to a record.

### 2015-07-24 doc/thesis/review.tex
Fixing glossary definitions and adding page referenes to them

The main goal of gradual typing \cite{siek2006gradual} is to allow
programmers to choose between static and dynamic typing in the same
language.
To do that, Siek and Taha \cite{siek2006gradual} extended the simply
typed lambda calculus with the dynamic type $?$, as we can see in
Figure \ref{fig:gtlc}.
In gradual typing, type annotations are optional, and an untyped
variable is syntactic sugar for a variable whose declared type is
the dynamic type $?$, that is, $\lambda x.e$ is equivalent to $\lambda x{:}?.e$.
Under these circumstances, we can view gradual typing as a way to add
a dynamic type to statically typed languages.

@@

The central idea of gradual typing is the \emph{consistency}
relation, written $T_{1} \sim T_{2}$.
The consistency relation allows implicit conversions to and from the
dynamic type, and disallows conversions between inconsistent types
\cite{siek2006gradual}.
For instance, $\Number \sim \;?$, $? \sim \Number$,
$\String \sim \;?$, and $? \sim \String$,
but $\Number \not\sim \String$, and
$\String \not\sim \Number$.
The consistency relation is both reflexive and symmetric, but
it is not transitive.

@@

Later, Siek and Taha \cite{siek2007objects} reported that the consistency relation
is orthogonal to the subtyping relation, so we can combine them to achieve
the \emph{consistent-subtyping} relation, written $T_{1} \lesssim T_{2}$.
This relation is essential for designing gradual type systems for
object-oriented languages.
Like the consistency relation, and unlike the subtyping relation,
the consistent-subtyping relation is not transitive.
Therefore, $\Number \lesssim \;?$, $? \lesssim \Number$,
$\String \lesssim \;?$, and $? \lesssim \String$,
but $\Number \not\lesssim \String$, and
$\String \not\lesssim \Number$.

>>>>

\label{def:dynamictype}
The main goal of gradual typing \cite{siek2006gradual} is to allow
programmers to choose between static and dynamic typing in the same
language.
To do that, Siek and Taha \cite{siek2006gradual} extended the simply
typed lambda calculus with the dynamic type $?$, as we can see in
Figure \ref{fig:gtlc}.
In gradual typing, type annotations are optional, and an untyped
variable is syntactic sugar for a variable whose declared type is
the dynamic type $?$, that is, $\lambda x.e$ is equivalent to $\lambda x{:}?.e$.
Under these circumstances, we can view gradual typing as a way to add
a dynamic type to statically typed languages.

@@

\label{def:consistency}
The central idea of gradual typing is the \emph{consistency}
relation, written $T_{1} \sim T_{2}$.
The consistency relation allows implicit conversions to and from the
dynamic type, and disallows conversions between inconsistent types
\cite{siek2006gradual}.
For instance, $\Number \sim \;?$, $? \sim \Number$,
$\String \sim \;?$, and $? \sim \String$,
but $\Number \not\sim \String$, and
$\String \not\sim \Number$.
The consistency relation is both reflexive and symmetric, but
it is not transitive.

@@

\label{def:consistent-subtyping}
Later, Siek and Taha \cite{siek2007objects} reported that the consistency relation
is orthogonal to the subtyping relation, so we can combine them to achieve
the \emph{consistent-subtyping} relation, written $T_{1} \lesssim T_{2}$.
This relation is essential for designing gradual type systems for
object-oriented languages.
Like the consistency relation, and unlike the subtyping relation,
the consistent-subtyping relation is not transitive.
Therefore, $\Number \lesssim \;?$, $? \lesssim \Number$,
$\String \lesssim \;?$, and $? \lesssim \String$,
but $\Number \not\lesssim \String$, and
$\String \not\lesssim \Number$.

### 2015-07-24 doc/thesis/system.tex
Fixing glossary definitions and adding page referenes to them

We also use the tags \emph{unique}, \emph{open}, \emph{fixed}, and \emph{closed}
to classify table types.
The tag \emph{unique} represents tables with no keys that do not
inhabit one of the table's key types, and with no alias.
In particular, the type of the table constructor has this tag.
The tag \emph{open} represents \emph{unique} table types that
have at least one alias.
The tag \emph{fixed} represents \emph{unique} table types that we
do not know how many aliases they have.
In particular, the type of a class has this tag.
The tag \emph{closed} represents table types that do not provide
any guarantees about keys with types not listed in the table type.
In particular, in the concrete syntax, type annotations, interface
declarations, and userdata declarations always describe \emph{closed} table types.
In the next sections we explain in more detail why we need
different table types.

@@

Typed Lua includes filter types as a way to discriminate the type of local
variables inside conditions.
Our type system uses filter types to formalize the \texttt{type} predicates
that we mentioned in Section \ref{sec:unions}.
This means that \texttt{type} predicates use filter types of the form
$\phi(T_{1},T_{2})$ to discriminate local variables that are bound to
union types.
In a filter type $\phi(T_{1},T_{2})$, $T_{1}$ is the original type and
$T_{2}$ is the discriminated type.

Typed Lua includes projection types as a way to project
unions of tuple types into unions of first-level types.
In Section \ref{sec:fap} we will show in more detail how our type system
uses them as a mechanism for handling unions of tuple types,
when they appear in the right-hand side of the declaration of local variables,
as we mentioned in Section \ref{sec:unions}.
We also show how this feature allows our type system to constrain
the type of a local variable that depends on the type of another local variable.

@@

The subtyping rule for pair types is the standard covariant rule:
\[
\begin{array}{c}
\mylabel{S-PAIR}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
       \senv \vdash P_{1} \subtype P_{2}}
      {\senv \vdash F_{1} \times P_{1} \subtype F_{2} \times P_{2}}
\end{array}
\]

@@

Filter types are subtypes only of themselves and of $\Value$.
More precisely, a filter type $\phi(F_{1},F_{2})$ is a subtype of
the same filter type $\phi(F_{1},F_{2})$, which shares the same
types $F_{1}$ and $F_{2}$, and it is also a subtype of $\Value$.

Projection types are subtypes only of themselves and of $\Value$.
More precisely, a projection type $\pi_{i}^{x}$ is a subtype of the
same projection type $\pi_{i}^{x}$, which shares the same union of
tuples $x$ and the same index $i$, and it is also a subtype of $\Value$.

@@

\section{Typing rules}
\label{sec:rules}

>>>>

\label{def:tabletype}
We also use the tags \emph{unique}, \emph{open}, \emph{fixed}, and \emph{closed}
to classify table types.
The tag \emph{unique} represents tables with no keys that do not
inhabit one of the table's key types, and with no alias.
In particular, the type of the table constructor has this tag.
The tag \emph{open} represents \emph{unique} table types that
have at least one alias.
The tag \emph{fixed} represents \emph{unique} table types that we
do not know how many aliases they have.
In particular, the type of a class has this tag.
The tag \emph{closed} represents table types that do not provide
any guarantees about keys with types not listed in the table type.
In particular, in the concrete syntax, type annotations, interface
declarations, and userdata declarations always describe \emph{closed} table types.
In the next sections we explain in more detail why we need
different table types.

@@

\label{def:filtertype}
Typed Lua includes filter types as a way to discriminate the type of local
variables inside conditions.
Our type system uses filter types to formalize the \texttt{type} predicates
that we mentioned in Section \ref{sec:unions}.
This means that \texttt{type} predicates use filter types of the form
$\phi(T_{1},T_{2})$ to discriminate local variables that are bound to
union types.
In a filter type $\phi(T_{1},T_{2})$, $T_{1}$ is the original type and
$T_{2}$ is the discriminated type.

\label{def:projectiontype}
Typed Lua includes projection types as a way to project
unions of tuple types into unions of first-level types.
In Section \ref{sec:fap} we will show in more detail how our type system
uses them as a mechanism for handling unions of tuple types,
when they appear in the right-hand side of the declaration of local variables,
as we mentioned in Section \ref{sec:unions}.
We also show how this feature allows our type system to constrain
the type of a local variable that depends on the type of another local variable.

@@

The subtyping rule for pair types is the standard covariant rule:
\[
\begin{array}{c}
\mylabel{S-PAIR1}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
       \senv \vdash P_{1} \subtype P_{2}}
      {\senv \vdash F_{1} \times P_{1} \subtype F_{2} \times P_{2}}
\end{array}
\]

@@

Filter types are subtypes only of themselves.
More precisely, a filter type $\phi(F_{1},F_{2})$ is a subtype of
the same filter type $\phi(F_{1},F_{2})$, which shares the same
types $F_{1}$ and $F_{2}$.

Projection types are subtypes only of themselves.
More precisely, a projection type $\pi_{i}^{x}$ is a subtype of the
same projection type $\pi_{i}^{x}$, which shares the same union of
tuples $x$ and the same index $i$.

The subtyping rules for expression list types are similar to the
subtyping rules for tuple types.

@@

\section{Type checking}
\label{sec:rules}
